---
title: "ODSMethods"
author: "Shawn Garbett, Lucy Bailu, Jonathan Schildcrout"
date: "2025-06-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ODSMethods)
```

## Introduction

When outcome covariates are available one can perform biased
sampling to maximize information returned, or outcome dependent sampling
(ODS). This is especially helpful
to minimize exposure or an expensive test. The statistical analysis
requires correction and this package has routines to assist in this
process via multiple methods. See 'Two-Phase, Generalized Case-Control
Designs for the Study of Quantitative Longitudinal Outcomes' Schildcrout,
et al. (2020) <doi: 10.1093/aje/kwz127>.

This process is similar to mixed effect modeling in that sampling is done
on an outcome that is a random effect, in the case of ascertainment corrected
maximum likelihood (ACML), the slope, intercept, mean or bivariate slope and intercept of a longitudinal response is used to inform sampling for the 
more expensive covariate. Cases in the tails of this random effect distribution are considered more informative and the "hump" in the middle
is less informative so sampling is biased.


## Design

### Simulated Dataset

The package includes a small simulated data set with known "truth". 

This is a simulated data set with known ``truth''. The dataset consists of
1000 patients and a measured response to a hypothetical drug over the course
of 5 months. Further, some ethnicity information is available if
Hispanic origin or not. An expensive genotype analysis exists, and to
minimize exposure a biased sample is performed. Patients that are in the
outer 20% of responses for intercept and slope are sampled for the genotyping
with a probability of 1. Patients that are in the inner 80% of responses for
intercept and slope are sampled with a probability of 0.25 for genotyping.
Thus on average 400 patients of the 1000 would be genotyped.

The model was generated using the following:

$$Y_{ij} = 10 + 0.5 \cdot Month - 0.5 \cdot Genotype  + 0.25 \cdot Month
\cdot Genotype + 0.5 \cdot Ancestry + b_{0i} + b_{1i} \cdot Month + \epsilon_{ij}$$

where

$$b_{0i} \sim \mathcal{N}(\mu=0, \sigma^2_0=4)$$
$$b_{1i} \sim \mathcal{N}(\mu=0, \sigma^2_1=0.25)$$
$$\rho(b_{0i}, b_{1i}) = 0.1$$
$$\epsilon_{ij} \sim \mathcal{N}(\mu=0, \sigma^2=1)$$

Further,

$$\pi(genotype_i = 1 | ancestry_i) = 0.1 + 0.15 \cdot I(ancestry_i=1)$$

```{r data}
data(gbti)
```

### Specify the Design

To specify the design, we'll using a simple intercept sampling model. The
design to sample assumes that the slope is the most informative. One
can plot the result of this design versus it's quantiles. 

```{r }
design <- ods(Response ~ Month|Patient, 'slope',
            p_sample=c(1, 0.25, 1),
            data=gbti, quantiles=c(0.1, 0.9))
summary(design)
plot(design)
```

## Sample

The design object contains a recommended sample for ACML.

```{r sampling_acml}
data <- gbti[gbti$Patient %in% design$sample_ids,]

```

## ACML

To fit the model using ascertainment corrected likelihood (ACML) a simple call
specifying the fixed effects. The random effects will be pulled from the
specified design. 

```{r acml, R.options=list(width=100)}
est <- acml(Response ~ Month*Genotype, design, data)
summary(est, digits=3)
```

One can also look at the coefficients transformed back to the original data scale.

```{r transform_back}
summary(est, digits=3, transform=TRUE)
```

## Visual summary

Let's look at a visual summary of the model.

```{r plot_acml}
#plot(est, which=1)
```

From this one can see, XXXX. 

Further, it's good to check the assumption that random effects are normal.

```{r plot_diagnostics}
#plot(est, which=2)
```

Finally, checking the residuals of the final model.
```{r plot_residuals}
#plot(est, which=3)
```


## References


