---
title: "ODSMethods Progress Report"
subtitle: "Package functionality and slope-sampling simulation performance"
author: "Lucy Yan"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    theme: cosmo
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)

pkg_root <- normalizePath(if (file.exists("DESCRIPTION")) "." else "..")
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(ODSMethods)
}

set.seed(20260519)
data(gbti)

sim_path <- file.path(pkg_root, "data", "mi_acml_summary_table_wide_b.csv")
sim_results <- read.csv(sim_path, check.names = FALSE, stringsAsFactors = FALSE)
sim_results <- sim_results[!is.na(sim_results$coef_name) & sim_results$coef_name != "", ]

coef_labels <- c(
  beta1 = "Intercept",
  beta2 = "Time",
  beta3 = "SNP",
  beta4 = "Confounder",
  beta5 = "SNP x time",
  beta6 = "log SD(intercept)",
  beta7 = "log SD(slope)",
  beta8 = "RE correlation",
  beta9 = "log residual SD"
)

sim_results$parameter <- ifelse(
  sim_results$coef_name %in% names(coef_labels),
  unname(coef_labels[sim_results$coef_name]),
  sim_results$coef_name
)

parse_percent <- function(x) {
  as.numeric(gsub("%", "", x, fixed = TRUE))
}

sim_results$rel_bias_acml_num <- parse_percent(sim_results$rel_bias_acml)
sim_results$rel_bias_mi_num <- parse_percent(sim_results$rel_bias_mi)
sim_results$rel_bias_se_acml_num <- parse_percent(sim_results$rel_bias_se_acml)
sim_results$rel_bias_se_mi_num <- parse_percent(sim_results$rel_bias_se_mi)

fmt_num <- function(x, digits = 3) {
  ifelse(is.na(x), NA, round(x, digits))
}

format_design_columns <- function(data, cols, n = 6, digits = 3) {
  out <- head(data[, cols, drop = FALSE], n)
  list_cols <- c("acml_samp_prob_i", "cutpoints_i")

  for (col in intersect(list_cols, names(out))) {
    if (is.list(out[[col]])) {
      out[[col]] <- vapply(
        out[[col]],
        function(x) paste(format(round(as.numeric(x), digits), trim = TRUE), collapse = ", "),
        character(1)
      )
    }
  }

  out
}
```

# Overview

This report summarizes the current progress of the `ODSMethods` R package.
The package is being developed for outcome-dependent sampling (ODS) methods in
two-phase study design for longitudinal data.

Here we demonstrate what the package can currently do with different input/design settings, and present estimate bias and standard error bias for ACML and multiple imputation (MI) under a slope-based ODS example.

# Package Motivation

Outcome-dependent sampling is useful when the longitudinal outcome is available
for a full phase-I cohort, but an exposure or biomarker is expensive
and can only be measured in a phase-II subset.

The package currently supports:

- constructing ODS and BDS two-phase sampling designs;
- visualizing the selected sampling regions;
- fitting ascertainment-corrected maximum likelihood (ACML) models;
- fitting weighted likelihood models.

# Example Phase-I Data

The package includes a simulated longitudinal dataset, `gbti`.

```{r}
str(gbti)
```

```{r}
data.frame(
  rows = nrow(gbti),
  subjects = length(unique(gbti$Patient)),
  visits_per_subject = length(unique(gbti$Month)),
  genotyped_subjects = length(unique(gbti$Patient[!is.na(gbti$Genotype)])),
  missing_genotype_rows = sum(is.na(gbti$Genotype))
)
```

The minimum design input is a long-format dataset with:

- a longitudinal response, here `Response`;
- a time variable, here `Month`;
- a subject identifier, here `Patient`;
- design choices such as sampling method, sampling probabilities, and cutpoints or quantiles.

```{r}
head(gbti)
```

# Design Formula

The design interface uses a formula of the form:

```r
Response ~ Month | Patient
```

This means:

- `Response`: the longitudinal outcome;
- `Month`: the time variable;
- `Patient`: the subject ID.

# Slope-Sampling Example

The main example uses slope sampling, where subjects with extreme estimated slopes (within subject variation) are sampled with higher probability.

```{r}
slope_design <- ods(
  Response ~ Month | Patient,
  method = "slope",
  p_sample = c(1, 0.25, 1),
  data = gbti,
  quantiles = c(0.10, 0.90)
)

summary(slope_design)
```

```{r}
#| fig-cap: "Slope-based ODS design. Points in the tails have higher sampling probability."
plot(slope_design)
```

The design object keeps the full phase-I dataset and adds design columns that
can be carried into the phase-II study.

```{r}
format_design_columns(slope_design$data, c(
  "Patient", "Month", "Response", "method_i",
  "acml_samp_prob_i", "cutpoints_i", "sampled"
))
```

```{r}
data.frame(
  total_subjects = length(unique(slope_design$data$Patient)),
  sampled_subjects = length(slope_design$sample_ids),
  sampled_rows = sum(slope_design$data$sampled == 1),
  observed_genotype_rows = sum(!is.na(slope_design$data$Genotype))
)
```

# BDS Example

```{r}
bds_slope_design <- bds(
  Response ~ Month | Patient,
  method = "slope",
  p_sample = c(1, 0.25, 1),
  data = gbti,
  quantiles = c(0.10, 0.90)
)

summary(bds_slope_design)
```

```{r}
#| fig-cap: "Slope-based BDS design."
plot(bds_slope_design)
```

```{r}
format_design_columns(bds_slope_design$data, c(
  "Patient", "Month", "Response", "method_i",
  "acml_samp_prob_i", "cutpoints_i", "sampled"
))
```

```{r}
data.frame(
  total_subjects = length(unique(bds_slope_design$data$Patient)),
  sampled_subjects = length(unique(bds_slope_design$data$Patient[bds_slope_design$data$sampled == 1])),
  sampled_rows = sum(bds_slope_design$data$sampled == 1)
)
```

# Complete Two-Phase Workflow

This section shows the workflow of a two-phase study. The first step is
done before the expensive covariate is measured. The second step is done after
the phase-II data collection is complete.

## Step 1: Choose The Study Design

Assume the phase-I clinical-trial data contain the longitudinal outcome and
basic variables, but not the expensive covariate. Here `Genotype` is removed to
mimic that setting.

```{r}
phase1_data <- gbti[, setdiff(names(gbti), "Genotype")]

phase1_design <- ods(
  Response ~ Month | Patient,
  method = "slope",
  p_sample = c(1, 0.25, 1),
  data = phase1_data,
  quantiles = c(0.10, 0.90)
)

summary(phase1_design)
```

The output suggests which subjects should be selected for phase-II
measurement.

```{r}
phase2_subjects <- unique(phase1_design$data$Patient[phase1_design$data$sampled == 1])

data.frame(
  phase1_subjects = length(unique(phase1_design$data$Patient)),
  selected_phase2_subjects = length(phase2_subjects),
  selected_phase2_rows = sum(phase1_design$data$sampled == 1)
)
```

The study team can export this subject list for lab measurements.

```{r}
head(data.frame(Patient = phase2_subjects))
```

## Step 2: Analyze The Completed Phase-II Data

After phase-II data collection, the expensive covariate is available for the
selected subjects. In this example, we create a completed lab result table for
the selected subjects.

```{r}
phase2_patient_info <- unique(
  gbti[gbti$Patient %in% phase2_subjects, c("Patient", "Ancestry", "Genotype")]
)
phase2_patient_info <- phase2_patient_info[!duplicated(phase2_patient_info$Patient), ]

set.seed(20260520)
missing_genotype <- is.na(phase2_patient_info$Genotype)
prob_genotype <- 0.10 + 0.15 * (phase2_patient_info$Ancestry == "Hispanic")
phase2_patient_info$Genotype[missing_genotype] <- as.logical(
  rbinom(sum(missing_genotype), 1, prob_genotype[missing_genotype])
)

phase2_lab_results <- phase2_patient_info[, c("Patient", "Genotype")]

analysis_data <- merge(
  phase1_design$data,
  phase2_lab_results,
  by = "Patient",
  all.x = TRUE
)

data.frame(
  analysis_rows = nrow(analysis_data),
  phase2_lab_subjects = nrow(phase2_lab_results),
  subjects_with_genotype = length(unique(analysis_data$Patient[!is.na(analysis_data$Genotype)])),
  rows_with_genotype = sum(!is.na(analysis_data$Genotype))
)
```

The design columns created before phase-II are kept in the analysis dataset:
`method_i`, `acml_samp_prob_i`, `cutpoints_i`, and `sampled`.

```{r}
format_design_columns(analysis_data, c(
  "Patient", "Month", "Response", "Genotype",
  "method_i", "acml_samp_prob_i", "cutpoints_i", "sampled"
))
```

Now rebuild the design object from the stored design columns and fit the ACML
analysis model.

```{r}
completed_design <- ods(
  Response ~ Month | Patient,
  data = analysis_data,
  method_name = "method_i",
  acml_samp_prob_name = "acml_samp_prob_i",
  cutpoints_name = "cutpoints_i",
  sampled_name = "sampled"
)

completed_fit <- acml(Response ~ Month * Genotype, completed_design)
summary(completed_fit, digits = 3, transform = TRUE)
```

This is the intended applied workflow: use `ods()` to define the phase-II
sample, store the design columns, then use the same design information for
the ACML analysis after data collection is complete.

# Other Design Inputs

The same interface can specify other ODS designs.

```{r}
intercept_design <- ods(
  Response ~ Month | Patient,
  method = "intercept",
  p_sample = c(1, 0.25, 1),
  data = gbti,
  quantiles = c(0.10, 0.90)
)

mean_design <- ods(
  Response ~ Month | Patient,
  method = "mean",
  p_sample = c(1, 0.25, 1),
  data = gbti,
  quantiles = c(0.10, 0.90)
)

bivariate_design <- ods(
  Response ~ Month | Patient,
  method = "bivariate",
  p_sample = c(0.25, 1),
  data = gbti,
  quantiles = 0.80
)

mixture_design <- ods(
  Response ~ Month | Patient,
  method = "mixture",
  p_sample = c(1, 0.25, 1),
  data = gbti,
  quantiles = c(0.10, 0.90),
  prob_intercept = 0.70
)

designs <- list(
  mean = mean_design,
  intercept = intercept_design,
  slope = slope_design,
  bivariate = bivariate_design,
  mixture = mixture_design
)

data.frame(
  method = names(designs),
  sampled_subjects = vapply(designs, function(x) length(x$sample_ids), integer(1)),
  sampled_rows = vapply(designs, function(x) sum(x$data$sampled == 1), integer(1))
)
```

```{r}
#| fig-cap: "Intercept-based ODS design."
plot(intercept_design)
```

```{r}
#| fig-cap: "Mean-response ODS design."
plot(mean_design)
```

```{r}
#| fig-cap: "Bivariate ODS design using both intercept and slope."
plot(bivariate_design)
```

```{r}
#| fig-cap: "Mixture ODS design using both intercept and slope."
plot(mixture_design)
```

# Quantiles and Cutpoints

Users can define sampling regions either by empirical quantiles:

```{r}
quantile_design <- ods(
  Response ~ Month | Patient,
  method = "slope",
  p_sample = c(1, 0.25, 1),
  data = gbti,
  quantiles = c(0.10, 0.90)
)

quantile_design$cutpoints
```

or by explicit cutpoints:

```{r}
cutpoint_design <- ods(
  Response ~ Month | Patient,
  method = "slope",
  p_sample = c(1, 0.25, 1),
  data = gbti,
  cutpoints = c(-0.20, 1.30)
)

cutpoint_design$cutpoints
```

# Analysis Example: ACML

After the design is specified, `acml()` fits the corrected longitudinal model.

```{r}
acml_fit <- acml(Response ~ Month * Genotype, slope_design)
summary(acml_fit, digits = 3, transform = TRUE)
```

```{r}
coef(acml_fit, transform = TRUE)
```

# Analysis Example: Weighted Likelihood

Weighted likelihood can use an existing sampling weight column.

```{r}
weighted_design <- ods(
  Response ~ Month | Patient,
  method = "intercept",
  p_sample = c(1, 0.25, 1),
  data = gbti,
  quantiles = c(0.10, 0.90),
  weights = "pSample"
)

wl_fit <- WL(Response ~ Month, weighted_design)
summary(wl_fit, digits = 3, transform = TRUE)
```

# Existing Phase-II Design Columns

If the sampling design has already been generated and stored in the dataset,
the package can rebuild a design object using those columns.

```{r}
rebuilt_design <- ods(
  Response ~ Month | Patient,
  data = slope_design$data,
  method_name = "method_i",
  acml_samp_prob_name = "acml_samp_prob_i",
  cutpoints_name = "cutpoints_i",
  sampled_name = "sampled"
)

data.frame(
  object_class = class(rebuilt_design)[1],
  total_subjects = length(unique(rebuilt_design$data$Patient)),
  sampled_subjects = length(unique(rebuilt_design$data$Patient[rebuilt_design$data$sampled == 1])),
  sampled_rows = sum(rebuilt_design$data$sampled == 1)
)

format_design_columns(rebuilt_design$data, c(
  "Patient", "Month", "Response", "method_i",
  "acml_samp_prob_i", "cutpoints_i", "sampled"
))
```

# Simulation Study: ACML and MI Bias

The simulation compares ACML and MI under slope sampling. For this report, the
focus is estimate bias (%) and standard error bias (%).

## Simulation Setting

The simulation setting follows a linear mixed model with slope-based phase-II sampling.

The two performance metrics shown here are:

- estimate bias (%), comparing the average estimate with the true parameter;
- SE bias (%), comparing the average model-based SE with the empirical SD.

## Estimate Bias

```{r}
estimate_bias_table <- data.frame(
  parameter = sim_results$parameter,
  true = fmt_num(sim_results$true_acml),
  acml_mean = fmt_num(sim_results$mean_est_acml),
  mi_mean = fmt_num(sim_results$mean_est_mi),
  estimate_bias_acml = sim_results$rel_bias_acml,
  estimate_bias_mi = sim_results$rel_bias_mi
)

knitr::kable(estimate_bias_table)
```

## Standard Error Bias

This table compares the empirical standard deviation with the average estimated
standard error.

```{r}
se_bias_table <- data.frame(
  parameter = sim_results$parameter,
  emp_sd_acml = fmt_num(sim_results$emp_sd_acml),
  mean_se_acml = fmt_num(sim_results$mean_se_acml),
  se_bias_acml = sim_results$rel_bias_se_acml,
  emp_sd_mi = fmt_num(sim_results$emp_sd_mi),
  mean_se_mi = fmt_num(sim_results$mean_se_mi),
  se_bias_mi = sim_results$rel_bias_se_mi
)

knitr::kable(se_bias_table)
```


# Input Checks

The package gives targeted messages for common input mistakes.

```{r}
show_error <- function(expr) {
  out <- try(expr, silent = TRUE)
  if (!inherits(out, "try-error")) return("No error")
  msg <- conditionMessage(attr(out, "condition"))
  msg <- gsub("[\r\n]+", " ", msg)
  msg <- gsub("\\*", "", msg)
  msg <- gsub("\\s+", " ", msg)
  trimws(msg)
}

input_checks <- data.frame(
  input_problem = c(
    "Missing | id in formula",
    "Unknown sampling method",
    "Wrong p_sample length for bivariate",
    "Both quantiles and cutpoints supplied"
  ),
  message = c(
    show_error(ods(Response ~ Month,
      method = "intercept", p_sample = c(1, 0.25, 1),
      data = gbti, quantiles = c(0.10, 0.90)
    )),
    show_error(ods(Response ~ Month | Patient,
      method = "random", p_sample = c(1, 0.25, 1),
      data = gbti, quantiles = c(0.10, 0.90)
    )),
    show_error(ods(Response ~ Month | Patient,
      method = "bivariate", p_sample = c(1, 0.25, 1),
      data = gbti, quantiles = 0.80
    )),
    show_error(ods(Response ~ Month | Patient,
      method = "slope", p_sample = c(1, 0.25, 1),
      data = gbti, quantiles = c(0.10, 0.90),
      cutpoints = c(-1, 1)
    ))
  )
)

knitr::kable(input_checks, format = "html")
```

# Current Status and Discussion Points

What is working now:

- ODS and BDS design construction for `mean`, `intercept`, `slope`, `bivariate`, and `mixture` design;
- design summaries and plots;
- ACML and weighted likelihood fitting;
- basic input validation.

Next steps:

- MI treating $X_e$ as ordinal, using `orm` function;
- what additional diagnostics should be added for `acml()` and `WL()` objects;
- Adding SMLE to the estimating functions.

# Reference

Schildcrout JS, Haneuse S, Tao R, Zelnick LR, Schisterman EF, Garbett SP,
Mercaldo ND, Rathouz PJ, Heagerty PJ. Two-Phase, Generalized Case-Control
Designs for the Study of Quantitative Longitudinal Outcomes. *American Journal
of Epidemiology*. 2020;189(2):81-90. doi: 10.1093/aje/kwz127.
